### 2018-01-27 submission/manuscript.tex
Look over edits after incorporating Ada suggestions

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness after correcting for study effect and
variable region sequenced with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in
relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
(P-value = 0.00034) while the reduction in Shannon diversity only
translated into an increased relative risk for developing carcinomas
(P-value = 0.0047). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

Although there has been an intense focus on microbiota-based biomarker
discovery for CRC, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with CRC. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some more
recent individual studies investigating the microbiota and CRC, making
it hard to know how extrapolatable these findings are. That particular
meta-analysis also added more potential microbial associations to both
carcinoma (\emph{Pantoea agglomerans} \emph{Ruminococcus},
\emph{Lactobacillus}) and adenoma (\emph{Prevotella},
\emph{Methanosphaera}, \emph{Succinovibrio}, \emph{Haemophilus
parainfluenzae}, \emph{Ruminococcus}, \emph{Lactobacillus}) stages of
CRC that need to be investigated further, since a number of these genera
have been found to be enriched in controls and not disease
{[}13,16,17{]}. There are also genera like \emph{Roseburia} that some
studies find to be increased in CRC but other studies report either
decreased or no difference {[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera. Groups
focusing on broad scale community metrics have found that metrics such
as richness are decreased in the adenoma stage of CRC versus controls.
Other studies have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. Based on
the studies mentioned, there seems to be very little overlap between the
genera identified to be associated with adenoma and carcinoma, with
\emph{Lactobacillus} being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk and the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not without risk of complications. Although stool
represents an easy and less invasive way to assess risk, it is not clear
how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue directly may
provide clearer answers but is not without limitations. The community
present for sampling following the colonoscopy bowel prep may reflect
the better adhered microbiota versus the resident community.
Additionally, these samples contain more host DNA, potentially limiting
the types of analysis that can be done. It is well known that low
biomass samples can be very difficult to work with and results can be
study dependent due to the randomness of contamination {[}29{]}.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Carcinomas:}} To assess differences in broad scale community metrics
as disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overal median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overal median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of CRC:}} Next we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered RR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in RR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. The top 5 most significantly increased RR were the same between
adenoma and carcinoma for both stool and colon tissue {[}Figure 3{]}.
Mouth-associated genera were commonly found in the top 5 genera
associated with an increased RR of having an adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased RR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased RR genera between both stages and sample site.

@@

In stool, the most common genera in the top 10 most imporant variables,
in models for the first approach, that used full genera-based models
were \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}
{[}Figure 6A \& B{]}. Regardless of sample type, mouth-associated genera
were present in models for the carcinoma stage of CRC {[}Figure 6A \&
B{]}. Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

@@

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, formation. The results that we
have presented show that the both genera and OTU select and full models,
for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes within the GI community is crucial
for carcinoma formation. Importantly, it suggests that there may be key
members of the GI community that might be studied further to potentially
reduce the risk of carcinoma. Conversely, using the present data, it is
clear that new approaches may be needed to identify members of the
community associated with adenoma stage of disease. Regardless of sample
type and whether a full or select model was used, our Random Forest
models consistently performed poorly. Yet, the step-wise decrease in
diversity suggests that the adenoma-associated community is not normal
but has changed subtly {[}Figure 1{]}. This change in diversity, at this
early stage of disease, could be focal to the adenoma itself. One
possible explanation is that how the host interacts with these subtle
changes at early stages of the disease is what leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both adenoma and
carcinoma {[}Figure 6{]}. These data suggest that whether the
non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with CRC Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S7B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of top 10 taxa being present only in a single
study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why the these results
seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated micobe, and the results from this study suggest that
regardless sample type, mouth-associated genera are commonly associated
with an increased RR for both adenoma and carcinoma stage of disease.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.11.1) package {[}39{]} and then Z-score
normalized using the car (v2.1.6) package {[}40{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.78) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.

>>>>

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness, after correcting for study effect and
variable region sequenced, with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in
relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
(P-value = 0.00034) while the reduction in Shannon diversity only
translated into an increased relative risk for developing carcinomas
(P-value = 0.0047). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

Although there has been an intense focus on microbiota-based biomarker
discovery for CRC, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with CRC. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some more
recent individual studies investigating the microbiota and CRC, making
it hard to know how extrapolatable these findings are. That particular
meta-analysis also added more potential microbial associations to both
carcinoma (\emph{Pantoea agglomerans} \emph{Ruminococcus},
\emph{Lactobacillus}) and adenoma (\emph{Prevotella},
\emph{Methanosphaera}, \emph{Succinovibrio}, \emph{Haemophilus
parainfluenzae}, \emph{Ruminococcus}, \emph{Lactobacillus}) stages of
CRC that need to be investigated further, since a number of these genera
have been found to be enriched in controls and not disease
{[}13,16,17{]}. Additionally, genera like \emph{Roseburia} have been
found in some studies to be increased in CRC but in others to either be
decreased or have no difference {[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera of
disease. Groups focusing on broad scale community metrics have found
that metrics such as richness are decreased in the adenoma stage of CRC
versus controls. Other studies have identified \emph{Lactococcus},
\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in
adenoma {[}25--27{]}. Based on the studies mentioned, there seems to be
very little overlap between the genera identified to be associated with
adenoma and carcinoma, with \emph{Lactobacillus} being one of the few
commonalities.

Targeting the identification of CRC microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk and the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not without risk of complications. Although stool
represents an easy and less invasive way to assess risk, it is not clear
how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue may provide clearer
answers but is not without limitations. The community present for
sampling following the colonoscopy bowel prep may reflect the better
adhered microbiota versus the resident community. Additionally, these
samples contain more host DNA, potentially limiting the types of
analysis that can be done. It is well known that low biomass samples can
be very difficult to work with and results can be study dependent due to
the randomness of contamination {[}29{]}.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of CRC:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overal median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of CRC:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered RR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in RR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased RR of having an adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased RR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased RR genera between both stages and sample site.

@@

In stool, the most common genera in the top 10 most imporant variables,
in the full community models using the first approach, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
6A \& B{]}. Regardless of sample type, mouth-associated genera were
present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

@@

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, formation. The results that we
have presented show that both the genera and OTU select and full models,
for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes within the GI community is crucial
for carcinoma formation. Importantly, it suggests that there may be key
members of the GI community that might be studied further to potentially
reduce the risk of carcinoma. Conversely, using the present data, it is
clear that new approaches may be needed to identify members of the
community associated with adenoma stage of disease. Regardless of sample
type and whether a full or select model was used, our Random Forest
models consistently performed poorly. Yet, the step-wise decrease in
diversity suggests that the adenoma-associated community is not normal
but has changed subtly {[}Figure 1{]}. This change in diversity, at this
early stage of disease, could be focal to the adenoma itself. One
possible explanation is that how the host interacts with these subtle
changes at early stages of the disease is what leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both adenoma and
carcinoma {[}Figure 6{]}. These data suggest that whether the
non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with CRC. Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S7B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of the top 10 taxa being present only in a
single study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why these results
seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated micobe, and the results from this study suggest that
regardless of sample type, mouth-associated genera are commonly
associated with an increased RR for both adenoma and carcinoma stage of
disease.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for \(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.78) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.
